MultifidelityTrajectoryOptimization

### 2024-11-18 artefacts/paper.tex
derive analytical form for correlation

\documentclass{article}
\usepackage[a4paper, total={7in, 8in}]{geometry}
\usepackage{graphicx} % Required for inserting images
\usepackage{amsmath} % Required for some math elements
\usepackage{hyperref} % Required for hyperlinks

@@

\section{Problem Formulation}
\subsection{Mass-Spring-Damper System}
For this paper, we consider a simplified trajectory optimization problem that is easy to understand and analyze.
We try to solve a reference tracking problem for the position and velocity of a simple mass-spring-damper system.
The mass-spring-damper system has the following parameters:
\begin{itemize}
  \item Mass: $m = 5$ g
  \item Spring constant: $k = 2$ N/m
  \item Damping coefficient: $c = 0.5$ Ns/m
    % \item Initial position: $x_0 = 0$ m
    % \item Initial velocity: $\dot{x}_0 = 0$ m/s
\end{itemize}

@@

We track the position $x$ and velocity $\dot{x}$ of the mass-spring-damper system using the control input $u$, which applies the force $F$ on the mass.
These are the states and control inputs of the system, giving use the following dynamics

@@

For a closed-loop system, we need to observe the output $y$ of the system, and in this case we observe the position $x$.

\begin{equation}
  y =
  \begin{bmatrix}
    1 & 0
  \end{bmatrix}
  \begin{bmatrix}
    x \\
    \dot{x}
  \end{bmatrix}
\end{equation}

This gives us an LTI system in the form of $\dot{x} = Ax + Bu$ and $y = Cx$, where

\begin{equation}
  \begin{aligned}
    A &=
    \begin{bmatrix}
      0 & 1 \\
      -\frac{k}{m} & -\frac{c}{m}
    \end{bmatrix} \\
    B &=
    \begin{bmatrix}
      0 \\
      \frac{1}{m}
    \end{bmatrix} \\
    C &=
    \begin{bmatrix}
      1 & 0
    \end{bmatrix}
  \end{aligned}
\end{equation}

\subsection{LQR Controller}
We set up an LQR controller to track the reference position and velocity of the mass-spring-damper system.
An LQR controller was chosen because it has a quadratic cost function that can be minimized analytically and gives us a
good reference for the trajectory optimization problem.

@@

\section{Control Variates}

@@

$u_h$ and $u_l$ each have $n_{u_h}$ and $n_{u_l}$ dimensions respectively, which represent the time steps of the simulation.
After converting them to the same dimensions as done in section \ref{comp_opt},
we can compute the cost for some realizations of the initial state $x_0$.
This will give us a set of costs of both high and low fidelity control inputs for a certain model/cost function.
We can then compute the correlation coefficient $\rho$ between the costs.
To illustrate this, assume we have $N$ realizations of the initial state $x_0$, and J is the cost function.
$u_h$ and $u_l$ are the control inputs for the high and low fidelity models respectively, where the dimensions have been
converted to the same number as described in section \ref{comp_opt}. $\mu_h$ and $\mu_l$ are the mean costs
of the high and low fidelity control inputs respectively, and $\sigma_h$ and $\sigma_l$ are the standard deviations of
the costs of the high and low fidelity control inputs respectively.

\begin{equation}
  \rho = \frac{1}{N-1} \sum_{i=1}^{N} \left( \frac{J(u_h, x_0^i) - \mu_h}{\sigma_h} \right) \left( \frac{J(u_l, x_0^i) - \mu_l}{\sigma_l} \right)
\end{equation}

@@

>>>>

\documentclass{article}
\usepackage[a4paper, total={7in, 8in}]{geometry}
\usepackage{graphicx} % Required for inserting images
\usepackage{amsmath} % Required for some math elements
\usepackage{hyperref} % Required for hyperlinks
\usepackage{amssymb} % Required for probability symbols

@@

\section{Problem Formulation}
This secition describes how the state space model of the dynamics is set up, along with the controller design and cost function for the trajectory optimization problem.
\subsection{Mass-Spring-Damper System}
For this paper, a simplified trajectory optimization problem is considered that is easy to understand and analyze.
The objective is to solve a reference tracking problem for the position and velocity of a simple mass-spring-damper system.
The mass-spring-damper system has the following parameters:
\begin{itemize}
  \item Mass: $m = 5$ g
  \item Spring constant: $k = 2$ N/m
  \item Damping coefficient: $c = 0.5$ Ns/m
    % \item Initial position: $x_0 = 0$ m
    % \item Initial velocity: $\dot{x}_0 = 0$ m/s
\end{itemize}

@@

The position $x$ and velocity $\dot{x}$ of the mass-spring-damper system can be changed using the control input $u$,
which applies the force $F$ on the mass. This yields the following dynamics

@@

For a closed-loop system, the output $y$ of the system needs to be observed, which is both the position and velocity in this case.

\begin{equation}
  y =
  \begin{bmatrix}
    1 & 0 \\
    0 & 1
  \end{bmatrix}
  \begin{bmatrix}
    x \\
    \dot{x}
  \end{bmatrix}
\end{equation}

This yields an LTI system in the form of $\dot{x} = Ax + Bu$ and $y = Cx$, where

\begin{equation}
  \begin{aligned}
    A &=
    \begin{bmatrix}
      0 & 1 \\
      -\frac{k}{m} & -\frac{c}{m}
    \end{bmatrix} \\
    B &=
    \begin{bmatrix}
      0 \\
      \frac{1}{m}
    \end{bmatrix} \\
    C &=
    \begin{bmatrix}
      1 & 0 \\
      0 & 1
    \end{bmatrix}
  \end{aligned}
\end{equation}

\subsection{LQR Controller} \label{lqr_ctrl}
We set up an LQR controller to track the reference position and velocity of the mass-spring-damper system.
An LQR controller was chosen because it has a quadratic cost function that can be minimized analytically and gives us a
good reference for the trajectory optimization problem.

@@

\subsection{Stochastic LQR Controller}
We will consider the case where the initial state $x_0$ is stochastic, with mean $\mu$ and covariance $\Sigma$.
The control input vector $U$ is still deterministic.

In this case, the state vector $X$ is stochastic, owing to the the initial condition $x_0$ being a vector of random variables. The state propagation equations are now expectations

\begin{equation}
  \begin{aligned}
    \mathbb{E}[x_1] &= \mathbb{E}[A x_0 + B u_0] = A \mathbb{E}[x_0] + B u_0 \\
    \mathbb{E}[x_2] &= \mathbb{E}[A x_1 + B u_1] = A \mathbb{E}[x_1] + B u_1 = A^2 \mathbb{E}[x_0] + A B u_0 + B u_1 \\
    \vdots \\
    \mathbb{E}[x_N] &= \mathbb{E}[A x_{N-1} + B u_{N-1}] = A \mathbb{E}[x_{N-1}] + B u_{N-1} \\
    &= A^N \mathbb{E}[x_0] + A^{N-1} B u_0 + A^{N-2} B u_1 + \ldots + B u_{N-1} \\
  \end{aligned}
\end{equation}

The matrices $S$ and $M$ remain the same, and we have the following expression for the expected value of the state vector $X$

$$
\mathbb{E}[X] = S U + M \mathbb{E}[x_0]
$$

The cost matrices $\bar{Q}$ and $\bar{R}$ are also the same. The cost function $J$ is now a random variable,
with an expected value $\mathbb{E}[J]$ and variance $Var[J]$.

The expected value of the cost function over all possible realizations of the initial conditions is
$$
\begin{aligned}
  \mathbb{E}[J] &= \mathbb{E}[x_N^T P x_N + \sum_{k=0}^{N-1} (x_k^T Q x_k + u_k^T R u_k)] \\
  &= \mathbb{E}[X^T \bar{Q} X + U^T \bar{R} U + x_0^T Q x_0] \\
  &= \mathbb{E}[X^T \bar{Q} X] + \mathbb{E}[U^T \bar{R} U] + \mathbb{E}[x_0^T Q x_0] \\
  &= \mathbb{E}[(S U + M x_0)^T \bar{Q} (S U + M x_0)] + U^T \bar{R} U + \mathbb{E}[x_0^T Q x_0] \\
  &= \mathbb{E}[U^T S^T \bar{Q} S U + U^T S^T \bar{Q} M x_0 + x_0^T M^T \bar{Q} S U + x_0^T M^T \bar{Q} M x_0] + U^T \bar{R} U + \mathbb{E}[x_0^T] Q \mathbb{E}[x_0] + \text{tr}(Q \text{Cov}[x_0]) \\
  &= U^T S^T \bar{Q} S U + \mathbb{E}[U^T S^T \bar{Q} M x_0] + \mathbb{E}[x_0^T M^T \bar{Q} S U] + \mathbb{E}[x_0^T M^T \bar{Q} M x_0] + U^T \bar{R} U + \mathbb{E}[x_0^T] Q \mathbb{E}[x_0] + \text{tr}(Q \text{Cov}[x_0]) \\
  &= U^T (S^T \bar{Q} S + \bar{R}) U + U^T S^T \bar{Q} M \mathbb{E}[x_0] + \mathbb{E}[x_0^T] M^T \bar{Q} S U + \mathbb{E}[x_0^T M^T \bar{Q} M x_0] + \mathbb{E}[x_0^T] Q \mathbb{E}[x_0] + \text{tr}(Q \text{Cov}[x_0]) \\
  &= U^T (S^T \bar{Q} S + \bar{R}) U + \mathbb{E}[x_0^T] M^T \bar{Q} S U + \mathbb{E}[x_0^T] M^T \bar{Q} S U + \mathbb{E}[x_0^T] M^T \bar{Q} M \mathbb{E}[x_0] + \text{tr}(M^T \bar{Q} M \text{Cov}[x_0]) \hdots \\
  & \quad + \mathbb{E}[x_0^T] Q \mathbb{E}[x_0] + \text{tr}(Q \text{Cov}[x_0]) \\
  &= U^T (S^T \bar{Q} S + \bar{R}) U + 2 \mathbb{E}[x_0^T] M^T \bar{Q} S U + \mathbb{E}[x_0^T] (M^T \bar{Q} M + Q) \mathbb{E}[x_0] + \text{tr}((M^T \bar{Q} M + Q) \text{Cov}[x_0]) \\
\end{aligned}
$$

Assuming $\mathbb{E}[x_0]$ and $\text{Cov}[x_0]$ is calculated using a Monte Carlo estimator, we can write the cost function as

$$
\begin{aligned}
  H &= S^T \bar{Q} S + \bar{R} = H^T \\
  q &= (\mathbb{E}[x_0^T] M^T \bar{Q} S)^T = S^T \bar{Q} M \mathbb{E}[x_0] \\
  c &= \mathbb{E}[x_0^T] (M^T \bar{Q} M + Q) \mathbb{E}[x_0] + \text{tr}((M^T \bar{Q} M + Q) \text{Cov}[x_0])
\end{aligned}
$$

The cost function can now be written as a quadratic function of the control inputs $U$

$$
J = U^T H U + 2 q^T U + c
$$

Which has a gradient with respect to $U$ as

$$
\Delta J = 2 H U + 2 q^T
$$

Assuming minimum exists, we can set the gradient to zero to get the optimal control input $U^*$

$$
\begin{aligned}
  \Delta J &= 2 H U + 2 q^T = 0 \\
  \implies U^* &= -H^{-1} q
\end{aligned}
$$

% ref for quadratic form: https://en.wikipedia.org/wiki/Quadratic_form_(statistics)

Tthe cost function can be rewritten as

$$
\begin{aligned}
  J &= U^T(S^T \bar{Q} S + \bar{R}) U + 2 x_0^T M^T \bar{Q} S U + x_0^T(M^T \bar{Q} M + Q) x_0 \\
  \text{Let}& \\
  K &= U^T(S^T \bar{Q} S + \bar{R}) U \\
  L &= 2 M^T \bar{Q} S U \\
  N &= M^T \bar{Q} M + Q = N^T \\
  &\text{(quadratic multiplication by diagnonal($\bar{Q}$) results in a symmetric, $Q$ is symmetric)} \\
  \text{Then}& \\
  J &= K + x_0^T L + x_0^T N x_0 \\
\end{aligned}
$$

Then, the expectation and variance of the cost function is as follows. Take note that the random variable here is $x_0$ and $\mathbb{E}[x_0]$ is a scalar.

% ref:
% - https://www.math.uwaterloo.ca/~hwolkowi/matrixcookbook.pdf
% - https://math.stackexchange.com/questions/2163694/expectation-of-quartic-form-for-multivariate-gaussian/4247240#4247240
% - https://math.stackexchange.com/questions/1302882/variance-of-a-quadratic-form

\begin{equation}
  \begin{aligned}
    \mathbb{E}[J] &= \mathbb{E}[K] + \mathbb{E}[x_0^T L] + \mathbb{E}[x_0^T N x_0] \\
    &= K + \mathbb{E}[x_0^T] L + \mathbb{E}[x_0^T] N \mathbb{E}[x_0] + \text{tr}(N \text{Cov}[x_0]) \\
    Var[J] &= Var[K + x_0^T L + x_0^T N x_0] \\
    &\text{note: $K$ is deterministic, so its variance is 0} \\
    & \text{note: $x_0^T L$ is a scalar, $x_0^T L = L^T x_0$} \\
    &= Var[L^T x_0 + x_0^T N x_0] \\
    &= 2\text{tr}(N \text{Cov}[x_0] N \text{Cov}[x_0]) + 4 \mathbb{E}[x_0^T]N\text{Cov}[x_0]N\mathbb{E}[x_0] + 4 L^T \text{Cov}[x_0] N \mathbb{E}[x_0] + L^T \text{Cov}[x_0] L \\
  \end{aligned}
\end{equation}

To determine the covariance between the costs of two solutions $U_1$ and $U_2$, the components of $J$ that depend on $U$
need to be written in terms of $U_1$ and $U_2$.

For $U_1$, the cost function is

$$
\begin{aligned}
  J(U_1) &= K_1 + x_0^T L_1 + x_0^T N x_0 \\
  \text{where} \\
  K_1 &= U_1^T(S^T \bar{Q} S + \bar{R}) U_1 \\
  L_1 &= 2 M^T \bar{Q} S U_1 \\
\end{aligned}
$$

Similarly, for $U_2$, the cost function is

$$
\begin{aligned}
  J(U_2) &= K_2 + x_0^T L_2 + x_0^T N x_0 \\
  \text{where} \\
  K_2 &= U_2^T(S^T \bar{Q} S + \bar{R}) U_2 \\
  L_2 &= 2 M^T \bar{Q} S U_2 \\
\end{aligned}
$$

% ref:
% - https://math.stackexchange.com/questions/1302882/variance-of-a-quadratic-form
% - https://stats.stackexchange.com/questions/303466/prove-that-mathrmcovxtax-xtbx-2-mathrmtra-sigma-b-sigma-4-mu
% - https://stats.stackexchange.com/questions/35084/covariance-of-a-linear-and-quadratic-form-of-a-multivariate-normal
% - https://www.researchgate.net/publication/243770190_Quadratic_Forms_in_Random_Variables P74, 3.2d.9
the covariance between their costs can be calculated as
\begin{equation}
  \begin{aligned}
    \text{Cov}[J(U_1), J(U_2)] &= \text{Cov}[K_1 + x_0^T L_1 + x_0^T N x_0, K_2 + x_0^T L_2 + x_0^T N x_0] \\
    &= \text{Cov}[x_0^T L_1 + x_0^T N x_0, x_0^T L_2 + x_0^T N x_0] \text{ (covariance is shift invariant)} \\
    &= \text{Cov}[x_0^T L_1, x_0^T L_2] + \text{Cov}[x_0^T L_1, x_0^T N x_0] + \text{Cov}[x_0^T N x_0, x_0^T L_2] + \text{Cov}[x_0^T N x_0, x_0^T N x_0] \\
    &= \text{Cov}[ L_1^T x_0, L_2^T x_0] + \text{Cov}[L_1^T x_0, x_0^T N x_0] + \text{Cov}[x_0^T N x_0, L_2^T x_0] + \text{Cov}[x_0^T N x_0, x_0^T N x_0] \\
    &= L_1^T \text{Cov}[x_0] L_2 + \text{Cov}[x_0^T N x_0, L_1^T x_0] + \text{Cov}[x_0^T N x_0, x_0^T L_2] + \text{var}[x_0^T N x_0] \\
    &= L_1^T \text{Cov}[x_0] L_2 + 2\mathbb{E}[x_0^T] N \text{Cov}[x_0] L_1 + 2\mathbb{E}[x_0^T] N \text{Cov}[x_0] L_2 \hdots \\
    & \quad 4 \mathbb{E}[x_0^T]N\text{Cov}[x_0]N\mathbb{E}[x_0] + 2\text{tr}((N \text{Cov}[x_0])^2) \\
  \end{aligned}
\end{equation}

\section{Multi Level Monte Carlo}

This section formulates the stochastic trajectory optimization problem as a multi-level Monte Carlo problem.

@@

$u_h$ and $u_l$ each have $n_{u_h}$ and $n_{u_l}$ dimensions respectively, which represent the time steps of the simulation.
After converting them to the same dimensions as done in section \ref{comp_opt},
the cost for some realizations of the initial state $x_0$ can be computed.
Hence, a set of costs of both high and low fidelity control inputs can be computed for a certain model/cost function.
The correlation coefficient $\rho$ can then be computed between the distribution of the cost function values.
To illustrate this, assume there are $N$ realizations of the initial state $x_0$, and $J$ is the cost function.
$u_h$ and $u_l$ are the control inputs for the high and low fidelity models respectively, where the dimensions have been
converted to the same number as described in section \ref{comp_opt}. $\mu_h$ and $\mu_l$ are the mean costs
of the high and low fidelity control inputs respectively, and $\sigma_h$ and $\sigma_l$ are the standard deviations of
the costs of the high and low fidelity control inputs respectively.

\begin{equation}
  \rho(J(u_h), J(u_l)) = \frac{1}{N-1} \sum_{i=1}^{N} \left( \frac{J(u_h, x_0^i) - \mu_h}{\sigma_h} \right) \left( \frac{J(u_l, x_0^i) - \mu_l}{\sigma_l} \right)
\end{equation}

This can be simplified to:

\begin{equation}
  \rho(J(u_h), J(u_l)) = \frac{\text{Cov}(J(u_h), J(u_l))}{\sigma_h \sigma_l}
\end{equation}

@@

The correlation can also be calculated analytically using the covariance of the costs in the high fidelity objective function and the low fidelity objective function.
Let

\begin{itemize}
  \item $J_h$: high fidelity cost function
  \item $U_1$: high fidelity control input
  \item $U_2$: upsampled low fidelity control input $u_{lh} = T_{lh} u_l$
  \item $J_1 = J_h(U_1, x_0)$
  \item $\sigma_1$: standard deviation of $J_1$, $\sqrt{\text{Var}[J_1]}$
  \item $\sigma_2$: standard deviation of $J_2$, $\sqrt{\text{Var}[J_2]}$
\end{itemize}

Then correlation coefficient between $J_1$ and $J_2$ is

\begin{equation}
  \begin{aligned}
    \rho(J_1, J_2) &= \frac{\text{Cov}(J_1, J_2)}{\sigma_1 \sigma_2} \\
    &= \frac{\text{Cov}(J_1, J_2)}{\sqrt{\text{Var}[J_1]} \sqrt{\text{Var}[J_2]}}
  \end{aligned}
\end{equation}

Where $\text{Cov}(J_1, J_2)$, $\text{Var}[J_1]$ and $\text{Var}[J_2]$ can be calculated using the eqations in section \ref{lqr_ctrl}.
